---
title: "Relationship of sex steroid hormone with lifestyle among US women"
author: "Ziqing Ye"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  

### Overview
This project investigates how sex steroid hormone vary due to demographics, BMI, smoking habits, alchohal comsuption, physical activities using the data of sex hormone markers retrieved from NHANES 2015-2016 Demographics Data, Examination Data, Labratory Data, and Questionnaire Data. Professors that I have talked to helped me with desgining the project and how to analyze the data.

[Link to my Github repository]: (https://github.com/ziqingye0210/BMIN503_Final_Project)


### Introduction 
Unhealthy sex hormone level in women may cause many diseases, such as Polycystic Ovary Syndrome (PCOS), androgen excess and cancer. An estimated 5 to 7 million women in the united States suffer with the effects of PCOS; and PCOS can occur in girls as young as 11 years old. Androgen excess in women and the associated chronic diseases are a public health concern. The data set retrieved from NHANES could be used to assist in prevention of diseases mentioned above.

To understand how sex steroid hormone vary due to demographics, BMI, smoking habits, alchohal comsuption, physical activities, the knowledge of what sex hormone markers could be used show changes in sex steroid hormone level is require. In addition, the bioinformatics knowledge that I learned in class also helped me find the approporate data set, retrieve data set, analyze data and use the result to address the research question raised. 


### Methods
The data sets used were retrieved from NANES 2015-2016 Demographics Data, Examination Data, Labratory Data, and Questionnaire Data. After retriving and cleaning data set, I applied linear regression models on each predictors and multivariate regresson models on all predictors to identify the preditors are most significantly related to change in sex steroid hormone level. Then I used k-means clustering to identify statistically determined profiles with sex hormone markers data. Finally, I analyzed women belong to which race, age range and BMI range are more likely to have higher or lower sex hormone levels.

> Download packages

```{r eval = TRUE}
# These two packages are used for determining the optimal number of k-mean clusters
pkgs <- c("factoextra",  "NbClust")
install.packages(pkgs)
```

> Retrieve data

```{r eval = TRUE}
library(nhanesA) 
library(plyr)
library(dplyr)
library(tidyverse)

# The data sets that I used are demographic variables and sample weights (DEMO_I), Sex Steroid Hormone - Serum (TST_I), Body Measures (BMX_I), Smoking - Cigarette Use (SMQ_I), Alcohol Use (ALQ_I), and Pysical Activity (PAQ_I).
varlist <- c("DEMO", "TST", "BMX", "SMQ", "ALQ", "PAQ") 
varlist_years <- paste0(varlist, c("_I"))

#Load everything from varlist_years in nhanes as a list
list_all <- sapply(varlist_years, function(x) {data.frame(nhanes(x))}) 

#Create a data.frame for each module
for(i in 1:length(list_all)) {
  assign(names(list_all)[i], list_all[[i]])
} 

#Combine years for each module
for (i in 1:length(varlist)){
  assign(varlist[i], plyr::rbind.fill(mget(grep(varlist[i], ls(), value = T))))
} 
rm(list = grep("_[DEFGHI]", ls(), value = T))

#Create a single data.frame that combines all modules 
nhanes.data <- merge(get(varlist[1]), get(varlist[2]), by = "SEQN", all = T)
for (i in 3:length(varlist)){
  nhanes.data <- merge(nhanes.data, get(varlist[i]), by="SEQN", all=T)
} 
rm(list = ls()[-which(ls() == "nhanes.data")])
```

> Clean data

```{r eval = TRUE}
#Choose varaibles
variables.wanted <- c("SEQN","RIAGENDR","RIDAGEYR","RIDRETH1", "LBXTST", "LBXEST", "LBXSHBG", "BMXBMI", "SMQ040", "ALQ120Q", "PAQ610", "PAQ620")

# Assign name to each variable and label levels for variables
# Exclude observations with no testosterone, estradiol, SHBG, BMI, activity data
nhanes <- nhanes.data %>%
  select(id=SEQN, gender=RIAGENDR, age=RIDAGEYR, race=RIDRETH1, testosterone=LBXTST, estradiol=LBXEST, SHBG=LBXSHBG, BMI=BMXBMI, smoke=SMQ040, alcohol=ALQ120Q, activity1=PAQ610, activity2=PAQ620) %>%
  mutate(gender=factor(gender, levels=c(1, 2), labels=c("male", "female"))) %>%
  mutate(race=factor(race, levels=c(3, 1, 2, 4, 5), labels=c("white", "MexicanAmerican", "Hispanic", "black", "other"))) %>%
  mutate(smoke=factor(smoke, levels=c(1, 2, 3), labels = c("EveryDay", "SomeDays", "NotAtAll"))) %>%
  filter(gender == "female") %>%
  filter(age %in% c(12:80)) %>%
  filter(!is.na(testosterone)) %>%
  filter(!is.na(estradiol)) %>%
  filter(!is.na(SHBG)) %>%
  filter(!is.na(BMI)) %>%
  mutate(BMI = cut(BMI, breaks=c(0, 18, 25, 30, 65), labels=c("Underweight", "NormalWeight", "Overweight", "Obese"))) %>%
  # activity 1 indicates number of days doing vigorous work and activity 2 indicates number of days doing moderate work
  filter(!is.na(activity1)|!is.na(activity2))

# Questionnaire was conducted among people over 18, therefore no smoking status reported for teenagers
# Exclude observations with age over 18 with no smoking status
# Smoking status for observations with age under 18 would be assigned to "NotAtAll"
nhanes <- filter(nhanes, age < 18 | (age >= 18 & !is.na(smoke)))
nhanes$smoke <- as.character(nhanes$smoke)
nhanes$smoke[is.na(nhanes$smoke)] <- "NotAtAll"
nhanes$smoke <- as.factor(nhanes$smoke)

# Questionnaire was conducted among people over 18, therefore no alcohol consumption reported for teenagers
# Exclude observations with age over 18 with no alcohol consumption status
# Alcohol consumption status for observations with age under 18 assigned to “0”, which means no alcohol consumption over the past 12 months
nhanes <- filter(nhanes, age < 18 | (age >= 18 & !is.na(alcohol)))
nhanes$alcohol <- as.numeric(nhanes$alcohol)
nhanes$alcohol[is.na(nhanes$alcohol)] <- 0
nhanes$alcohol <- cut(nhanes$alcohol, breaks=c(-Inf, 0, 3, 6, Inf), labels=c("0", "1-3", "4-6", "6+"))
nhanes$alcohol <- as.factor(nhanes$alcohol) 

# Divide age into five categories
nhanes$age <- cut(nhanes$age, breaks=c(11, 18, 29, 49, 69, Inf), labels=c("12-18", "19-29", "39-49", "50-69", "70+"))

# Get the total number of days doing moderate to vigorous activities
# Observations with total number of days doing moderate to vigorous activities over 7 days will be set to 7
# Delete column activity1 and column activity2
nhanes$activity <- rowSums(nhanes[,c("activity1", "activity2")], na.rm=TRUE)
nhanes$activity[nhanes$activity > 7] <- 7
nhanes <- nhanes[, -c(11:12)]

# The overall information of participants in this questionnaire, including gender, age, race, testosterone level, estradiol level, SHBG level, BMI level, smoking status, alcohol consumption and physical activity
summary(nhanes)

```

> Exploratory analysis

```{r eval = TRUE}
library(ggplot2)
library(cowplot)

# Box plots and corresponding linear regression results for age, race, BMI, smoke, alcohol, and activity
g1 <- ggplot(data = nhanes, aes(x = age, y = testosterone)) + geom_boxplot() + ylim(c(1, 200))
g2 <- ggplot(data = nhanes, aes(x = age, y = estradiol)) + geom_boxplot() + ylim(c(1, 2000))
g3 <- ggplot(data = nhanes, aes(x = age, y = SHBG)) + geom_boxplot() + ylim(c(1, 400))

plot_grid(g1, g2, g3, labels = "AUTO")

summary(lm(testosterone ~ age, data = nhanes))
summary(lm(estradiol ~ age, data = nhanes))
summary(lm(SHBG ~ age, data = nhanes))


r1 <- ggplot(data = nhanes, aes(x = race, y = testosterone)) + geom_boxplot() + ylim(c(1, 200))
r2 <- ggplot(data = nhanes, aes(x = race, y = estradiol)) + geom_boxplot() + ylim(c(1, 2000))
r3 <- ggplot(data = nhanes, aes(x = race, y = SHBG)) + geom_boxplot() + ylim(c(1, 400))

plot_grid(r1, r2, r3, labels = "AUTO")

summary(lm(testosterone ~ race, data = nhanes))
summary(lm(estradiol ~ race, data = nhanes))
summary(lm(SHBG ~ race, data = nhanes)) 


b1 <- ggplot(data = nhanes, aes(x = BMI, y = testosterone)) + geom_boxplot() + ylim(c(1, 200))
b2 <- ggplot(data = nhanes, aes(x = BMI, y = estradiol)) + geom_boxplot() + ylim(c(1, 2000))
b3 <- ggplot(data = nhanes, aes(x = BMI, y = SHBG)) + geom_boxplot() + ylim(c(1, 400))

plot_grid(b1, b2, b3, labels = "AUTO")

summary(lm(testosterone ~ BMI, data = nhanes))
summary(lm(estradiol ~ BMI, data = nhanes))
summary(lm(SHBG ~ BMI, data = nhanes))


s1 <- ggplot(data = nhanes, aes(x = smoke, y = testosterone)) + geom_boxplot() + ylim(c(1, 200))
s2 <- ggplot(data = nhanes, aes(x = smoke, y = estradiol)) + geom_boxplot() + ylim(c(1, 2000))
s3 <- ggplot(data = nhanes, aes(x = smoke, y = SHBG)) + geom_boxplot() + ylim(c(1, 400))

plot_grid(s1, s2, s3, labels = "AUTO")

summary(lm(testosterone ~ smoke, data = nhanes))
summary(lm(estradiol ~ smoke, data = nhanes))
summary(lm(SHBG ~ smoke, data = nhanes))


h1 <- ggplot(data = nhanes, aes(x = alcohol, y = testosterone)) + geom_boxplot() + ylim(c(1, 200))
h2 <- ggplot(data = nhanes, aes(x = alcohol, y = estradiol)) + geom_boxplot() + ylim(c(1, 2000))
h3 <- ggplot(data = nhanes, aes(x = alcohol, y = SHBG)) + geom_boxplot() + ylim(c(1, 400))

plot_grid(h1, h2, h3, labels = "AUTO")

summary(lm(testosterone ~ alcohol, data = nhanes))
summary(lm(estradiol ~ alcohol, data = nhanes))
summary(lm(SHBG ~ alcohol, data = nhanes))


a1 <- ggplot(data = nhanes, aes(x = factor(activity), y = testosterone)) + geom_boxplot() + ylim(c(1, 200))
a2 <- ggplot(data = nhanes, aes(x = factor(activity), y = estradiol)) + geom_boxplot() + ylim(c(1, 2000))
a3 <- ggplot(data = nhanes, aes(x = factor(activity), y = SHBG)) + geom_boxplot() + ylim(c(1, 400))

plot_grid(a1, a2, a3, labels = "AUTO")

summary(lm(testosterone ~ activity, data = nhanes))
summary(lm(estradiol ~ activity, data = nhanes))
summary(lm(SHBG ~ activity, data = nhanes))
```

### Results

```{r eval = TRUE}
library(ggbiplot)
library(factoextra)
library(NbClust)

# Significant predictors are age and BMI, even when adjusting for other variables.
summary(lm(testosterone ~ age + BMI, data = nhanes))
summary(lm(estradiol ~ age + BMI, data = nhanes))
summary(lm(SHBG ~ age + BMI, data = nhanes))

summary(lm(testosterone ~ age + race + BMI + smoke + alcohol + activity, data = nhanes))
summary(lm(estradiol ~ age + race + BMI + smoke + alcohol + activity, data = nhanes))
summary(lm(SHBG ~ age + race + BMI + smoke + alcohol + activity, data = nhanes))

# Only keep columns of age, testosterone level, estradiol level, SHBG level and BMI
nhanes.cluster <- nhanes[c(3, 5:8)]
str(nhanes.cluster)


# funcion of blom transformation
blom = function(x, method="general", alpha=pi/8, 
                complete=FALSE, na.last="keep", na.rm=TRUE,
                adjustN=TRUE,
                min=1, max=10, ...){
  if(complete){x=x[complete.cases(x)]}
  Ranks = rank(x, na.last=na.last, ...)
  if(adjustN==FALSE){N = length(x)}
  if(adjustN==TRUE) {N = sum(complete.cases(x))}
  if(method=="blom")   {Score = qnorm((Ranks-0.375)/(N+0.25))}
  if(method=="vdw")    {Score = qnorm((Ranks)/(N+1))}
  if(method=="tukey")  {Score = qnorm((Ranks-1/3)/(N+1/3))}
  if(method=="rankit") {Score = qnorm((Ranks-1/2)/(N))}
  if(method=="elfving"){Score = qnorm((Ranks-pi/8)/(N-pi/4+1))}
  if(method=="general"){Score = qnorm((Ranks-alpha)/(N-2*alpha+1))}
  if(method=="zscore") {Score = (x-mean(x, na.rm=na.rm))/sd(x, na.rm=na.rm)}
  if(method=="scale")  {
    Score = (((x - min(x, na.rm=na.rm)) * 
            (max - min)) / 
            (max(x, na.rm=na.rm) - min(x, na.rm=na.rm)) + 
             min)
  }
  return(Score)
}

# Use blom transformation to make sex hormone marker variables unit-free, so that they are comparable in cluster analysis
nhanes.cluster$testosterone <- blom(nhanes.cluster$testosterone, method = "blom")
nhanes.cluster$estradiol <- blom(nhanes.cluster$estradiol, method = "blom")
nhanes.cluster$SHBG <- blom(nhanes.cluster$SHBG, method = "blom")


# The result of pca indicates that the data are not simply grouped into clusters based on one single variable
nhanes.pca <- prcomp(nhanes.cluster[ , 2:4], scale = TRUE)
print(nhanes.pca)
ggbiplot(nhanes.pca, groups = nhanes.cluster$BMI, circle = TRUE, ellipse = TRUE) +
    scale_color_discrete(name = '') +
    theme(legend.direction = 'horizontal', legend.position = 'top')

ggbiplot(nhanes.pca, groups = nhanes.cluster$age, circle = TRUE, ellipse = TRUE) +
    scale_color_discrete(name = '') +
    theme(legend.direction = 'horizontal', legend.position = 'top')


# Use Elbow method to determine the optimal number of clusters for k-mean clustering
fviz_nbclust(nhanes.cluster[, 2:4], kmeans, method = "wss") +
    geom_vline(xintercept = 4, linetype = 2)+
  labs(subtitle = "Elbow method")


# Perform k-mean clustering using sex steroid hormone markers
nhanes.kmeans <- kmeans(nhanes.cluster[, 2:4], 4)
nhanes.kmeans


# Plot 2 dimensions at a time to interprete patterns
# SHBG is most helpful with seperating clusters
centers <- as.data.frame(nhanes.kmeans$centers)

ggplot(data = nhanes.cluster, aes(testosterone, SHBG, color = factor(nhanes.kmeans$cluster))) +
    geom_point() +
    geom_point(data = centers, aes(testosterone, SHBG), color = "purple", size = 3)

ggplot(data = nhanes.cluster, aes(testosterone, estradiol, color = factor(nhanes.kmeans$cluster))) +
    geom_point() +
    geom_point(data = centers, aes(testosterone, estradiol), color = "purple", size = 3)

ggplot(data = nhanes.cluster, aes(estradiol, SHBG, color = factor(nhanes.kmeans$cluster))) +
    geom_point() +
    geom_point(data = centers, aes(estradiol, SHBG), color = "purple", size = 3)


# Assign cluster number to each observation
nhanes <- cbind(nhanes, clusterNum = nhanes.kmeans$cluster)
nhanes <- nhanes[c(3, 5:8, 12)]
  

# Analyze each cluster for the following:
# Number of people from each age group, race or BMI range
# Mean of testoserone, estradiol, and SHBG
nhanes %>%
  group_by(clusterNum) %>%
  summarise(testosterone = mean(testosterone), estradiol = mean(estradiol), SHBG = mean(SHBG))
# The results shows the following:
# Observations in cluster 1 have the highest mean estradiol level and the highest mean SHBG level.
# Observations in cluster 2 have the lowest mean testosterone level and the lowest mean SHBG level.
# Observations in cluster 3 have the lowest mean estradiol level.
# Observations in cluster 4 have the highest testosterone level.

# The majority of observasions in cluster 1 are in age range of 12-18 and 39-49 and their weight are normal.
cluster1 <- subset(nhanes, clusterNum == 1)
cluster1 <- cluster1[c(1:5)]
summary(cluster1)

# The majority of observasions in cluster 2 are in age range of 50-69 and are obese.
cluster2 <- subset(nhanes, clusterNum == 2)
cluster2 <- cluster2[c(1:5)]
summary(cluster2)

# The majority of observasions in cluster 3 are in age range of 50-69.
cluster3 <- subset(nhanes, clusterNum == 3)
cluster3 <- cluster3[c(1:5)]
summary(cluster3)

# The majority of observasions in cluster 4 are in age range of 12-18 and are obese.
cluster4 <- subset(nhanes, clusterNum == 4)
cluster4 <- cluster4[c(1:5)]
summary(cluster4)

```

> Conclusiion

In conclusion, there is no significant correlation found between race, smoking habits, alchohal comsuption, physical activities and sex steroid hormone levels among US women. However, age and BMI are found to be significant predictors. Each sex steroid hormone marker is more likely to be associated with one or more age or BMI groups of observations. Although predictors that are more directly indicating lifestyle are not significant in our case, BMI could be treated as an indirect indicator of lifestyle. 

Based on the data set chosen and the analysis of the data, obese teenage girls are more likely to have higher testosterone level. Obese women in the age range of 50-69 are more likely to have low testosterone and SHBG level. Teenage girls and women in the age range of 39-49 with normal weight are more likely to have higher estradiol and SHBG level. Women in the age range of 50-69 tend to have lower estradiol level. Both teenage girls and women in ange range from 39-69 are more likely to have diseases associated with unhealthy level of sex hormone and obesity increases the possibility for sex hormone related diseases. 
