---
title: "Relationship of sex sterid hormone with lifestyle in US women"
author: "Ziqing Ye"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***
Use this template to complete your project throughout the course. Your Final Project presentation in class will be based on the contents of this document. Replace the title/name and text below with your own, but keep the headers.

### Overview
In this section, give a brief a description of your project and its goal, what data you are using to complete it, and what three faculty/staff in different fields you have spoken to about your project with a brief summary of what you learned from each person. Include a link to your final project GitHub repository.

> How does sex hormone vary due to demographics, weight(BMI), and diet/fasting habit (possibly smoking habits and alchohal comsuption) 

> include physical activity

> use mapping??

> what concentration of sex hormone would lead to diseases?


### Introduction 
In the first paragraph, describe the problem addressed, its significance, and some background to motivate the problem.

In the second paragraph, explain why your problem is interdisciplinary, what fields can contribute to its understanding, and incorporate background related to what you learned from meeting with faculty/staff.

> 

### Methods
In the first paragraph, describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. 

> Retrieve data

```{r eval = TRUE}
library(nhanesA) 
library(plyr)
library(dplyr)

varlist <- c("DEMO", "TST", "BMX", "FASTQX", "SMQ", "ALQ") 
varlist_years <- paste0(varlist, c("_I"))

#Load everything from varlist_years in nhanes as a list
list_all <- sapply(varlist_years, function(x) {data.frame(nhanes(x))}) 

#Create a data.frame for each module
for(i in 1:length(list_all)) {
  assign(names(list_all)[i], list_all[[i]])
} 

#Combine years for each module
for (i in 1:length(varlist)){
  assign(varlist[i], plyr::rbind.fill(mget(grep(varlist[i], ls(), value = T))))
} 
rm(list = grep("_[DEFGHI]", ls(), value = T))

#Create a single data.frame that combines all modules 
nhanes.data <- merge(get(varlist[1]), get(varlist[2]), by = "SEQN", all = T)
#The following for loop could be used if more than two modules selected
for (i in 3:length(varlist)){
  nhanes.data <- merge(nhanes.data, get(varlist[i]), by="SEQN", all=T)
} 
rm(list = ls()[-which(ls() == "nhanes.data")])

```

> Clean data

```{r eval = TRUE}
#Choose varaibles
variables.wanted <- c("SEQN","RIAGENDR","RIDAGEYR","RIDRETH1", "LBXTST", "LBXEST", "LBXSHBG", "BMXBMI", "SMQ040", "ALQ130", "PHAFSTHR")

#Assign name to each variable
nhanes.practice <- nhanes.data %>%
    select(id=SEQN, gender=RIAGENDR, age=RIDAGEYR, race=RIDRETH1, testosterone=LBXTST, estradiol=LBXEST, SHBG=LBXSHBG, BMI=BMXBMI, smoke=SMQ040, alcohol=ALQ130, fast=PHAFSTHR) %>%
    mutate(gender=factor(gender, levels=c(1, 2), labels=c("male", "female"))) %>%
    mutate(race=factor(race, levels=c(3, 1, 2, 4, 5), labels=c("white", "MexicanAmerican", "Hispanic", "black", "other"))) %>%
    mutate(smoke=factor(smoke, levels=c(1, 2, 3, 7, 9), labels = c("EveryDay", "SomeDays", "NotAtAll", "Refused", "DontKnow")))
```

> Exploratory analysis

```{r eval = TRUE}
library(ggplot2)
library(cowplot)

```

### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.
