---
title: "Relationship of sex sterid hormone with lifestyle among US women"
author: "Ziqing Ye"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***
Use this template to complete your project throughout the course. Your Final Project presentation in class will be based on the contents of this document. Replace the title/name and text below with your own, but keep the headers.

### Overview
In this section, give a brief a description of your project and its goal, what data you are using to complete it, and what three faculty/staff in different fields you have spoken to about your project with a brief summary of what you learned from each person. Include a link to your final project GitHub repository.

> How does sex hormone vary due to demographics, weight(BMI), smoking habits, alchohal comsuption, physical activities

### Introduction 
In the first paragraph, describe the problem addressed, its significance, and some background to motivate the problem.

In the second paragraph, explain why your problem is interdisciplinary, what fields can contribute to its understanding, and incorporate background related to what you learned from meeting with faculty/staff.


### Methods
In the first paragraph, describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. 

> Retrieve data

```{r eval = TRUE}
library(nhanesA) 
library(plyr)
library(dplyr)
library(tidyverse)

varlist <- c("DEMO", "TST", "BMX", "SMQ", "ALQ", "PAQ") 
varlist_years <- paste0(varlist, c("_I"))

#Load everything from varlist_years in nhanes as a list
list_all <- sapply(varlist_years, function(x) {data.frame(nhanes(x))}) 

#Create a data.frame for each module
for(i in 1:length(list_all)) {
  assign(names(list_all)[i], list_all[[i]])
} 

#Combine years for each module
for (i in 1:length(varlist)){
  assign(varlist[i], plyr::rbind.fill(mget(grep(varlist[i], ls(), value = T))))
} 
rm(list = grep("_[DEFGHI]", ls(), value = T))

#Create a single data.frame that combines all modules 
nhanes.data <- merge(get(varlist[1]), get(varlist[2]), by = "SEQN", all = T)
for (i in 3:length(varlist)){
  nhanes.data <- merge(nhanes.data, get(varlist[i]), by="SEQN", all=T)
} 
rm(list = ls()[-which(ls() == "nhanes.data")])

```

> Clean data

```{r eval = TRUE}
#Choose varaibles
variables.wanted <- c("SEQN","RIAGENDR","RIDAGEYR","RIDRETH1", "LBXTST", "LBXEST", "LBXSHBG", "BMXBMI", "SMQ040", "ALQ120Q", "PAQ610", "PAQ620")

#Assign name to each variable
#Get rid of useless observations
nhanes <- nhanes.data %>%
  select(id=SEQN, gender=RIAGENDR, age=RIDAGEYR, race=RIDRETH1, testosterone=LBXTST, estradiol=LBXEST, SHBG=LBXSHBG, BMI=BMXBMI, smoke=SMQ040, alcohol=ALQ120Q, activity1=PAQ610, activity2=PAQ620) %>%
  mutate(gender=factor(gender, levels=c(1, 2), labels=c("male", "female"))) %>%
  mutate(race=factor(race, levels=c(3, 1, 2, 4, 5), labels=c("white", "MexicanAmerican", "Hispanic", "black", "other"))) %>%
  mutate(smoke=factor(smoke, levels=c(1, 2, 3), labels = c("EveryDay", "SomeDays", "NotAtAll"))) %>%
  filter(gender == "female") %>%
  filter(age %in% c(12:80)) %>%
  filter(!is.na(testosterone)) %>%
  filter(!is.na(estradiol)) %>%
  filter(!is.na(SHBG)) %>%
  filter(!is.na(BMI)) %>%
  mutate(BMI = cut(BMI, breaks=c(0, 18, 25, 30, 65), labels=c("Underweight", "NormalWeight", "Overweight", "Obese"))) %>%
  filter(!is.na(activity1)|!is.na(activity2))

# Questionnaire was conducted among people over 18
# Exclude observations with age over 18 with no smoking status
# Smoking status for observations with age under 18 would be assigned to "NotAtAll"
nhanes <- filter(nhanes, age < 18 | (age >= 18 & !is.na(smoke)))
nhanes$smoke <- as.character(nhanes$smoke)
nhanes$smoke[is.na(nhanes$smoke)] <- "NotAtAll"
nhanes$smoke <- as.factor(nhanes$smoke)

# Questionnaire was conducted among people over 18
# Exclude observations with age over 18 with no alcohol consumption status
# Alcohol consumption status for observations with age under 18 assigned to “0”, which means no alcohol over the past 12 months
nhanes <- filter(nhanes, age < 18 | (age >= 18 & !is.na(alcohol)))
nhanes$alcohol <- as.numeric(nhanes$alcohol)
nhanes$alcohol[is.na(nhanes$alcohol)] <- 0
nhanes$alcohol <- cut(nhanes$alcohol, breaks=c(-Inf, 0, 3, 6, Inf), labels=c("0", "1-3", "4-6", "6+"))
nhanes$alcohol <- as.factor(nhanes$alcohol) 


# Get the total number of days doing moderate to vigorous activities
# Observations with days of activities larger than 7 days will be excluded
# Delete column activity1 and column activity2
nhanes$activity <- rowSums(nhanes[,c("activity1", "activity2")], na.rm=TRUE)
nhanes <- filter(nhanes, activity < 8)
nhanes <- nhanes[, -c(11:12)]

# The overall information of participants in this questionnaire, including gender, age, race, testosterone level, estradiol level, SHBG level, BMI level, smoking status, alcohol consumption and physical activity
summary(nhanes)



```

> Exploratory analysis

```{r eval = TRUE}
library(ggplot2)
library(cowplot)

# Scatter plot for age and each sex hormone marker expression level
ggplot(data = nhanes, aes(x = age, y = testosterone)) +
    geom_point()

# Box plot for race, BMI, smoke, alcohol, and activity
r1 <- ggplot(data = nhanes, aes(x = factor(race), y = testosterone)) + geom_boxplot()
r2 <- ggplot(data = nhanes, aes(x = factor(race), y = estradiol)) + geom_boxplot()
r3 <- ggplot(data = nhanes, aes(x = factor(race), y = SHBG)) + geom_boxplot()

plot_grid(r1, r2, r3, labels = "AUTO")


b1 <- ggplot(data = nhanes, aes(x = factor(BMI), y = testosterone)) + geom_boxplot()
b2 <- ggplot(data = nhanes, aes(x = factor(BMI), y = estradiol)) + geom_boxplot()
b3 <- ggplot(data = nhanes, aes(x = factor(BMI), y = SHBG)) + geom_boxplot()

plot_grid(b1, b2, b3, labels = "AUTO")

s1 <- ggplot(data = nhanes, aes(x = factor(smoke), y = testosterone)) + geom_boxplot()
s2 <- ggplot(data = nhanes, aes(x = factor(smoke), y = estradiol)) + geom_boxplot()
s3 <- ggplot(data = nhanes, aes(x = factor(smoke), y = SHBG)) + geom_boxplot()

plot_grid(s1, s2, s3, labels = "AUTO")


h1 <- ggplot(data = nhanes, aes(x = factor(alcohol), y = testosterone)) + geom_boxplot()
h2 <- ggplot(data = nhanes, aes(x = factor(alcohol), y = estradiol)) + geom_boxplot()
h3 <- ggplot(data = nhanes, aes(x = factor(alcohol), y = SHBG)) + geom_boxplot()

plot_grid(h1, h2, h3, labels = "AUTO")

a1 <- ggplot(data = nhanes, aes(x = factor(activity), y = testosterone)) + geom_boxplot()
a2 <- ggplot(data = nhanes, aes(x = factor(activity), y = estradiol)) + geom_boxplot()
a3 <- ggplot(data = nhanes, aes(x = factor(activity), y = SHBG)) + geom_boxplot()

plot_grid(a1, a2, a3, labels = "AUTO")
```

### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.

```{r eval = TRUE}

```
