---
title: "Effects of Intranasal Insulin on Cognitive and Function Performance in Pakinson Disease"
author: "Ziqing Ye"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***

### Overview
In this section, give a brief a description of your project and its goal, what data you are using to complete it, and what three faculty/staff in different fields you have spoken to about your project with a brief summary of what you learned from each person. Include a link to your final project GitHub repository.

>Insulin has been suggested to be a potential treatment for cognitive impairment in Parkinson's disease. The goal of this project is to determine whether intranasal insulin can affect cognition and moter performance in Parkinson's disease based on the double-blinded placebo-controlled pilot study data. The data that I am using is from "Novak, V., Novak, P. (2019). Safety and Preliminary Efficacy of Intranasal Insulin for Cognitive Impairment in Parkinson Disease and Multiple System Atrophy. PhysioNet".

### Introduction 
In the first paragraph, describe the problem addressed, its significance, and some background to motivate the problem.
In the second paragraph, explain why your problem is interdisciplinary, what fields can contribute to its understanding, and incorporate background related to what you learned from meeting with faculty/staff.
>Insulin is a peptide hormone that is used for treating diabetes. It was also found to be a potential treatment for cognitive impairment in Parkinson's disease. Because the brain area affected in Parkinson's disease was found to have insulin receptors and the insulin receptor signaling tend to decrease in Parkinson's disease. Taking intranasal insulin could help avoid hypoglycemia or insulin resistance due to the increasing insulin levels.
>To do such analysis, knowledge of Parkinson's disease, how insulin level relates to Parkinson's disease and how the data could be used to prove the efficacy of insulin in treating Parkinson's disease are needed. Also, how to analyze data with statistic knowledge is important to this project.

### Methods
In the first paragraph, describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. 
```{r eval = FALSE}

```

### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.
```{r eval = FALSE}

```